---
title: "coursera"
author: "Jochen Schmiedbauer"
date: "Sunday, November 16, 2014"
output: html_document
---
    Coursera course Reproducible Research Project Assignment 1
======================================
    ## Setting Knitr options
    
```{r setoptions }
library(knitr)
opts_chunk$set(fig.path = "figures/" )

```
    
    
    ## Loading and preprocessing the data
    
    The data is stored in a csv file and the working directory is set to the file location.

```{r loaddata}
setwd("D:/Temp/coursera/reprod_research") ## make sure that the file is present in this directory.
data <- read.csv("activity.csv")
data$date <- as.Date(data$date, format = "%Y-%m-%d")

```

## Histogram

Make a histogram of the total number of steps taken each day.

```{r 1_Histogram_steps_per_day, fig.width = 12}
library("ggplot2")
dailySteps <- aggregate(steps ~ date, data, sum)
ggplot(data,aes(date)) + geom_histogram(aes(weight=steps),binwidth=1)

```

## Mean and Median Steps per day

Calculate and report the mean and median total number of steps taken per day.

```{r Descriptive mean}
dMean <- mean(dailySteps$steps)
```

The mean is `r dMean`

```{r Descriptive median}
dMedian <- median(dailySteps$steps)

```

The median is `r dMedian`

## Mean steps per interval

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).

```{r 2_Interval_Graph_wo_NA, fig.width = 12}
mInterval <- aggregate(steps ~ interval, data, mean)
ggplot(mInterval, aes(interval,steps)) + geom_line()

```

## 5 minute interval with max number of average steps

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r Max}
maxR <- which.max(mInterval$steps)
maxI <- mInterval$interval[maxR]
maxS <- mInterval$steps[maxR]
```
The max interval average is `r maxI` with `r maxS` steps.

##Missing Values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
missing <- sum(is.na(data$steps))
```
There are `r missing` missing values. (NA rows)

## Replace NA's with Mean steps per interval

using the mean number of steps per interval as replacement for the missing values.

```{r replaceErrors}
oldData <- data
fn <- function(steps,interval) ifelse(is.na(steps),mInterval$steps[mInterval$interval==interval],steps)
data$steps <- apply(data[,c("steps","interval")],1, function(y) fn(y["steps"],y["interval"]) )

```
##Histogram with nulls replaced by interval averages

```{r 3_Histogram_steps_per_day_including_NA, fig.width=12}

dailySteps2 <- aggregate(steps ~ date, data, sum)
ggplot(data,aes(date)) + geom_histogram(aes(weight=steps),binwidth=1)

```
##New Mean and Median

```{r Descriptive Mean}
dMean2 <- mean(dailySteps2$steps)
```

The mean is `r dMean2`

```{r Descriptive Median}
dMedian2 <- median(dailySteps2$steps)
```

The mean is `r dMedian2`

##Change in Mean and Median

```{r BeforeAfter}
MeanDiff = dMean2 - dMean
MedianDiff = dMedian2 - dMedian

```
Impact of filling missing values:
    The mean changed by `r MeanDiff` steps and 
    the median changed by `r MedianDiff` steps.

##Days of the week

```{r}
dowFn <- function(day) ifelse(is.na(match(weekdays(day), c("Saturday","Sunday"))),"Weekday","Weekend")
data$dow <- factor(apply(data["date"],1,function(y) dowFn(as.Date(y["date"]))))

```

##Compute average interval steps for weekends and weekdays

```{r}
weekend <- data[data$dow=="Weekend",]
weekday <- data[data$dow=="Weekday",]

wkdMean <- aggregate(steps ~ interval, weekday, mean)
wkdMean$dow <- "Weekday"
wkendMean <- aggregate(steps ~ interval, weekend, mean)
wkendMean$dow <- "Weekend"
CombMeans <- rbind(wkdMean,wkendMean)
```

##Weekday versus weekend interval averages

```{r 4_weekday_versus_weekend, fig.width=12}
library(lattice)
xyplot(steps ~ interval|dow, data = CombMeans, layout = c(1,2), type = "l")

```